Loop-Aware Optimizations in PyPy's Tracing JIT

### 2012-08-17 talk/dls2012/paper.tex
PyPy's tracing JIT -> RPython's tracing JIT
don't insist on the pdf as a source
more correct explanation for the sqrt(Fix16) behaviour. Thanks, Mike.

Mike Pall pioneered a solution to address this problem in the context of a
dynamic language using a tracing JIT compiler. He published his algorithm and
its rationale in 2009~\cite{pall_luajit_2009} and implemented it in LuaJIT
2.0\footnote{\texttt{http://luajit.org/}}, an open source JIT compiler for the Lua
language. His approach allows to reuse all forward pass
optimizations to achieve loop invariant code motion and other loop-related
optimizations, which greatly simplifies the implementation. We have implemented
the same approach in PyPy's tracing JIT compiler, the results of which we
present here.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{benchmarks/result.pdf}
\end{center}
\caption{Benchmark results normalized to the runtime of the C version. The CPython results have been omitted to make the plot readable.}
\label{fig:benchmarks_plot}
\end{figure}

@@

The slowdown of sqrt(Fix16) compared to sqrt(int) or sqrt(float) show the
overhead of using a custom class with operator overloading for arithmetic. For
C/C++, this overhead is very low, for CPython the code becomes 30 times slower.
In LuaJIT, the overhead is a slowdown of 70\%. For PyPy, sqrt(Fix16) is only
slightly slower than sqrt(int), which is itself three times slower than
sqrt(float). This is probably due to the additional overflow checking necessary
for integer arithmetic in Python. The fact that LuaJIT and PyPy do so well on
sqrt(Fix16) shows that the allocation removal/sinking optimizations work well
in both JITs.

>>>>

Mike Pall pioneered a solution to address this problem in the context of a
dynamic language using a tracing JIT compiler. He published his algorithm and
its rationale in 2009~\cite{pall_luajit_2009} and implemented it in LuaJIT
2.0\footnote{\texttt{http://luajit.org/}}, an open source JIT compiler for the Lua
language. His approach allows to reuse all forward pass
optimizations to achieve loop invariant code motion and other loop-related
optimizations, which greatly simplifies the implementation. We have implemented
the same approach in RPython's tracing JIT compiler, the results of which we
present here.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{benchmarks/result}
\end{center}
\caption{Benchmark results normalized to the runtime of the C version. The CPython results have been omitted to make the plot readable.}
\label{fig:benchmarks_plot}
\end{figure}

@@

The performance of sqrt(Fix16) compared to the C version gives an indication of the
overhead of using a custom class with operator overloading for arithmetic. For
CPython the overhead over C is a lot larger than that of sqrt(int).
In LuaJIT, the overhead is very small. For PyPy, sqrt(Fix16) 2.2 times slower
than the C version. However, that is not actually due to the overhead of
operator overloading but due to the additional overflow checking necessary
for integer arithmetic in Python. The JIT does not manage to prove that the
integer operations in these benchmarks cannot overflow and therefore cannot
optimize away the overflow checking. This is also the reason why sqrt(float) is
so much faster than sqrt(int) for PyPy.
The fact that LuaJIT and PyPy do so well on
sqrt(Fix16) shows that the allocation removal/sinking optimizations work well
in both JITs.
